---
title: "Loading Eyelink data with eyelinker"
author: "Simon Barthelmé"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading Eyelink data with eyelinker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We'll use test data supplied by SR Research (which I found in the cili package for Python). The test data can be found in the extdata/ directory of the package.

```{r}
require(eyelinker)
#Look for file 
fpath <- system.file("extdata/mono500.asc.gz",package="eyelinker")
```

asc files can be gigantic, so it's a good idea to compress them, R doesn't mind (here they're compressed in gzip format, hence the .gz).

To read the file just call read.asc:

```{r}
dat <- read.asc(fpath)
```

dat is a list with fields:

```{r}
names(dat)
```

- raw is the raw data (eye position, velocity, etc.) as a function of time
- sac are the saccade events as labelled by the Eyelink
- fix are the fixations
- blinks are the blinks
- msg are message events
- info contains some meta-data

## Meta-data

Some meta-data can be read from the "SAMPLES" lines in the asc file.

```{r}
str(dat$info)
```

- velocity: true if data contains eye velocity 
- resolution: true if data contains resolution
- cr: true if corneal reflection mode is used
- htarg: true if data contains remote info (only applicable in remote setup)
- input: true if data contains input info
- left: true if left eye is recorded
- right: true if right eye is recorded
- mono: true if recording is monocular

Here we have a monocular recording of the left eye. 

## Raw data

The raw data has a simple structure:

```{r}
raw <- dat$raw
head(raw,3)
```

- time is a time stamp (ms)
- xp, yp:  x and y position of the recorded eye
- ps: pupil size (arb. units)
- cr.info: status of corneal reflection tracking.  "..." means all's well. See manual for more.

The block variable is probably irrelevant in most cases (the .asc file is divided into START and END blocks, and the block variable indexes them).

In a binocular recording the raw data has the following structure:

```{r}
dat.bi <- system.file("extdata/bino1000.asc.gz",package="eyelinker") %>% read.asc
head(dat.bi$raw,3)
```

The variables are the same as before, with the addition of a postfix corresponding to the eye (i.e. xpl is the x position of the left eye).



