---
title: "Loading Eyelink data with eyelinker"
author: "Simon Barthelmé"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading Eyelink data with eyelinker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We'll use test data supplied by SR Research (which I found in the cili package for Python). The test data can be found in the extdata/ directory of the package.

```{r}
require(eyelinker)
require(dplyr)

#Look for file 
fpath <- system.file("extdata/mono500.asc.gz",package="eyelinker")
```

asc files can be gigantic, so it's a good idea to compress them, R doesn't mind (here they're compressed in gzip format, hence the .gz).

To read the file just call read.asc:

```{r}
dat <- read.asc(fpath)
```

dat is a list with fields:

```{r}
names(dat)
```

- raw is the raw data (eye position, velocity, etc.) as a function of time
- sac are the saccade events as labelled by the Eyelink
- fix are the fixations
- blinks are the blinks
- msg are message events
- info contains some meta-data

## Meta-data

Some meta-data can be read from the "SAMPLES" lines in the asc file.

```{r}
str(dat$info)
```

- velocity: true if data contains eye velocity 
- resolution: true if data contains resolution
- cr: true if corneal reflection mode is used
- htarg: true if data contains remote info (only applicable in remote setup)
- input: true if data contains input info
- left: true if left eye is recorded
- right: true if right eye is recorded
- mono: true if recording is monocular

Here we have a monocular recording of the left eye. 

## What are the units? 

Depending on how the Eyelink is set up, positions can be reported in pixels or degrees, relative to the head, the screen or the camera.
I'm guessing the most common case is to use screen coordinates, but I don't know whether the coordinate system is stored in a predictable manner in asc files. If you have any suggestions please email me.
I'll assume you know what the relevant units are. 

## Raw data

The raw data has a simple structure:

```{r}
raw <- dat$raw
head(raw,3)
```

- time is a time stamp (ms)
- xp, yp:  x and y position of the recorded eye
- ps: pupil size (arb. units)
- cr.info: status of corneal reflection tracking.  "..." means all's well. See manual for more.
- block: the .asc file is divided into START and END blocks, and the block variable indexes them.

In a binocular recording the raw data has the following structure:

```{r}
dat.bi <- system.file("extdata/bino1000.asc.gz",package="eyelinker") %>% read.asc

head(dat.bi$raw,3)
```

The variables are the same as before, with the addition of a postfix corresponding to the eye (i.e. xpl is the x position of the left eye).


## Tidying up raw data

It's sometimes more convenient for plotting and analysis if the raw data are in "long" rather than "wide" format, as in the following example:

```{r}
library(tidyr)

raw.long <- dplyr::select(raw,time,xp,yp,block) %>% gather("coord","pos",xp,yp)
head(raw.long,2)
tail(raw.long,2)
```

The eye position is now in a single column rather than two, and the column "coord" tells us if the valuye corresponds to the x or y position. The benefits may not be obvious now, but it does make plotting the traces via ggplot2 a lot easier:


```{r}
require(ggplot2)
raw.long <- mutate(raw.long,ts=(time-min(time))/1e3) #let's have time in sec. 
ggplot(raw.long,aes(ts,pos,col=coord))+geom_point()
```

In this particular file there are four separate recording periods, corresponding to different "blocks" in the asc file, which we can check using:

```{r}
ggplot(raw.long,aes(ts,pos,col=coord))+geom_line()+facet_wrap(~ block)
```

## Saccades

The Eyelink automatically detects saccades in an online fashion. The results are converted to a data.frame: 

```{r}
sac <- dat$sac
head(sac,2)
```

Each line corresponds to a saccade, and the different columns are:

- stime and etime: the start and end times of the saccade
- dur: duration (ms)
- sxp, yxp: starting position
- exp, eyp: end position
- ampl: saccade amplitude
- pv: peak velocity
- block: see above

In the binocular case, we have:

```{r}
head(dat.bi$sac,3)
```

The only difference is in the "eye" column, which tells you in which eye the saccade was first recorded. 

## Labelling the saccades in the raw traces

To see if the saccades have been labelled correctly, we'll have to find the corresponding time samples in the raw data.

```{r}


```
